\textsc{\wobble: a data-driven method for precision radial velocities}

### 2018-12-20 paper/paper.tex
fitting 51peg with exoplanet

\documentclass[modern]{aastex62}

@@

\begin{abstract}
% Context
Extreme-precision radial velocity (\EPRV) exoplanet surveys face considerable data analysis challenges in extracting maximally precise \RVs from spectra. 
Chief among these, particularly for the upcoming generation of red-optimized spectrographs targeting M dwarfs, is the presence of telluric absorption features which are not perfectly known. 
Another major limitation on the achievable \RV precision is the need to adopt an imperfect stellar template against which to cross-correlate or otherwise match the observations. 
In both cases, the precision-limiting reliance on external information can be sidestepped using the data itself. 
% Aims
Here we propose a data-driven method to simultaneously extract precise \RVs and infer the underlying stellar and telluric spectra using a linear model (in the log of flux). 
The model employs a convex objective and convex regularization to keep the optimization of the spectral components fast. 
We implement this method in \wobble, an open-source \python package which uses \TF in one of its first non-neural-network applications to astronomical data. 
In this work, we demonstrate the performance of \wobble on archival \HARPS spectra.
% Results
We recover the canonical exoplanet 51 Pegasi b, detect the secular \RV evolution of the M dwarf \Mdwarf, and retrieve the \RM for the Hot Jupiter HD 189733b. 
We also present extremely high-\SNR composite stellar spectra and detailed time-variable telluric spectra derived from these data. 
\end{abstract}

@@

Data analysis pipelines for absorption cell instruments have long used a stellar template that is customized to the star in question, although traditionally this template is a single observation taken for the purpose \citep[e.g.][]{Butler1996}. 
Higher quality templates can be derived iteratively from the data \citep{Sato2002, Gao2016}. 
Such data-driven templates have the benefit of being both customized to the star in question and high-\SNR without requiring observing overhead. 
%\todo{(explain how this differs from what we'll do - v similar in spirit although ours is a full forward model with stellar spectrum as a latent variable)}

@@

Here we develop a data-driven model to infer the telluric and stellar spectra and calculate the stellar \RV at each observed epoch. 
The telluric model component may vary with time in a low-dimensional manner, which is also inferred from the data. 
Our model requires no prior knowledge of the star or the Earth's atmosphere. 
As such, it does not yield absolute measurements of \RVs, only highly precise relative measurements between epochs.

@@

In Section \ref{s:methods}, we outline the model and its key underlying assumptions. 
We present an open-source implementation of this method in \python and \TF
\citep{Abadi15} called \wobble.
In Section \ref{s:results}, we demonstrate \wobble's capabilities by applying our method to \HARPS archival data for three target stars: the canonical planet-hosting solar analog 51 Peg, the quiet M dwarf \Mdwarf, and the Hot Jupiter host HD 189733. 
We look further into the detailed time-variable telluric spectra inferred from these data in Section \ref{s:tellurics}.
We revisit many of the assumptions underlying \wobble in Section \ref{s:future} with a detailed look at the limitations of the current implementation. 
We outline potential ways of adapting \wobble\ for such cases as instruments with absorption cells, intrinsic time variability in the stellar spectrum, and lower-quality data. 
Finally, we conclude with a brief summary in Section \ref{s:conclusion}.

@@

For practical applications, the number of epochs $N$ must be large ($N \gg N_{components}$). 
This is a fundamental restriction rooted in the fact that every spectrum being modeled introduces a large number of free parameters to the model. 
In most cases, these epochs will also need to span a significant fraction of the observing season. 
In order to disentangle the stellar and telluric spectral components, the spectra must undergo Doppler shifts with respect to each other that are at least as large as a resolution element of the spectrograph. 
For stars which do not physically undergo large \RV shifts over very short timescales, this means that it is necessary to observe the star over a significant fraction of the year to take advantage of the changes in the Earth's projected motion. 
We will comment on ways to overcome this restriction in Section \ref{s:future}.

@@

We take the data to be the $M \times N$ matrix Y, where each entry $y_{m,n}$ is the observed logarithmic flux $\ln f$ for pixel $m$ of $M$ at epoch $n$ of $N$. 
We also have a corresponding $M \times N$ matrix of wavelength solutions. 
Again, the entries of this matrix are in logarithmic form ($\ln \lambda$). 
This choice makes the Doppler shift an additive term, considerably simplifying the math to follow.

@@

The stellar spectrum contribution is: 
\begin{equation}
y_{n}^\star = P(v_{obs, n}) \mu_{\star},
\label{eqn:star}
\end{equation}
where $\mu_{\star}$ is a spectral template with length $M'$ corresponding to
the length of an arbitrarily-chosen wavelength grid of control points, and
$P$ is a linear operator which applies a Doppler shift by velocity $v_n$ and
interpolates $\mu_\star$ into the observatory rest frame.
Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
\begin{equation}
 \ln \lambda(v) = \ln \lambda_{0} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
\end{equation}
The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space.
$P$ could be represented as a rectangular $M \times M'$ sparse matrix, but the dot product can be implemented efficiently using linear interpolation instead of instantiating the full matrix.

The apparent stellar \RV, $v_{obs, n}$, is a sum of the star's actual velocity about its center of mass ($v_{\star, n}$) and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV), the latter of which is known.

The telluric spectrum contribution is:
\begin{equation}
y_{n}^t =  A_n(\mu_{t} + W_{t} z_n).
\end{equation}
In addition to its mean spectrum $\mu_t$, the telluric component also includes a time-dependent term assembled from two variables: $W_t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$. 
$W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of basis vectors used. 
For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications. 
The net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $A_n$, a known quantity.

@@

L1 normalization adds a term to the log-likelihood that takes the form:
\begin{equation}
\Lone{\lambda}{p} \equiv -\lambda \sum_{i} | p_{i} | ,
\end{equation}
where $p$ is the vector of parameters to be normalized (in this case $\mu_{{\star}}$, $\mu_{{t}}$, or $W_{t}$) and $\lambda$ is the regularization amplitude.

@@

With regularization included, the final model likelihood to be optimized is:
\begin{equation}
\begin{split}
\ln \mathcal{L} = & -\frac{1}{2} \sum_{n} (y_{n} - y'_{n})^T C_{n}^{-1} (y_{n}-y'_{n})  + \Lone{\lambda_1}{\mu_{\star}} + \Ltwo{\lambda_2}{\mu_{\star}} \\
 &  + \Lone{\lambda_3}{\mu_{t}} + \Ltwo{\lambda_4}{\mu_{t}}  + \Lone{\lambda_5}{W_t}  + \Ltwo{\lambda_6}{W_t} + \Lone{1.0}{z_t} ,
\end{split}
\end{equation}
%\todo{BEDELL: THIS IS WRONG. We were supposed to set $\lambda_5$ to unity. My (Hogg's) bad. However, I wrote the following to justify our WRONG choice!}
where the regularization amplitude on the basis weights is arbitrarily set to unity, and all other regularization amplitudes $\lambda$ are set by grid searches using the above-described validation procedure.
The basis-weight regularization amplitude can be set arbitrarily to unity because there is a perfect degeneracy between amplifying the basis vectors and attenuating the basis weights.
This regularization choice breaks that perfect degeneracy.
The basis weights are regularized to encourage our desired outcome where the mean telluric spectrum contains as much telluric information as possible, while the variable basis picks up only the necessary time-variable changes.

@@

As an initial guess, we set the star to be stationary, e.g. $v_{obs} = $ \BERV at all epochs. 
We may then initialize the stellar spectrum $\mu_{\star}$ by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength $\lambda'$. 
The telluric template is initialized similarly by using the residuals after the stellar contribution has been removed, again binning in model wavelength (this time without applying any Doppler shift to the spectra) and taking the median values of each bin. 
Finally, we initialize the telluric spectrum's basis vectors $W_t$ and weights $z_t$ by performing Principal Component Analysis (\acronym{PCA}) on the residuals after both the stellar spectrum and the mean telluric spectrum have been removed. 
The $K$ highest eigenweights and their corresponding eigenvectors are taken as the basis weights and vectors.

Once all parameters are initialized, we optimize them iteratively. 
The likelihood function is maximized first by varying the stellar and telluric templates (including the telluric basis vectors). 
This step is a convex optimization, meaning that a global optimum for the template parameters should be reached under the condition of fixed time-dependent variables. 
Next, the templates are held fixed and the time-dependent parameters (stellar \RVs and telluric basis weights) are varied to maximize the likelihood function again. 
Technically the velocities are location parameters and their optimization is not convex.
However, once the velocities are known to a small fraction of a pixel (which they usually are, in practice, in advance of running \wobble), the linearized problem becomes convex at each iteration.
We repeat this procedure, optimizing spectra and time-dependent parameters in turn, until the likelihood appears to converge, typically within 100 iterations.
This iterative procedure is equivalent to a full simultaneous optimization in the limit of many iteration to convergence.

@@

The above-described model can be implemented in a variety of ways. We chose to build our code, \wobble, in \code{python} using \TF. 
\TF is a model building framework that has been primarily designed for machine
learning applications, but at its core \TF is fundamentally a collection of
highly optimized routines for doing linear algebra and efficiently computing the
derivatives of these models with respect to large numbers of parameters.
\wobble can be represented in this framework so we have re-purposed \TF for our
needs.
By using this framework, we benefit from the high performance and scalability
of the implementation, as well as the algorithms implemented within \TF for
fitting large numbers of parameters to large datasets.
%While it has primarily been used in the context of deep learning in the astronomical literature, its auto-differentiation and fast linear algebra capabilities make it valuable for this application as well.
The necessary optimizations over many parameters can be performed with high efficiency by \TF: the below-described analysis of 91 \HARPS spectra of 51 Peg, including the optimization of 72 spectral orders, runs in 60 minutes on a standard Mac desktop.

@@

@@

Despite the sparse phase coverage available across the planetary orbit, we recover a signal at the expected period and semi-amplitude in the \RVs (Figure \ref{fig:51peg_planet}). 
We fit a Keplerian signal with six free parameters (period $P$, \RV semi-amplitude $K$, eccentricity $e$, argument of periastron $\omega$, time of periastron $T_p$, and \RV offset $c$) using a simple $\chi^2$ optimization with \code{scipy.optimize.leastsq}. 
The resulting best-fit parameters for the \wobble \RVs are consistent with the literature values: $P = 4.2308$ d, $K = 55.9 \ms$, $e = 0.05$

Fits to an individual spectrum are shown in Figure \ref{fig:51peg_spectrum}. 
We emphasize that no \textit{a priori} information on e.g. expected spectral line positions and depths were used. 
The current realization of the \wobble algorithm treats each control point of the template spectra as independent and has no line shape parameterization included. 
Nevertheless, the optimized stellar templates clearly reproduce the expected appearance of the spectra in a variety of regimes, from a crowded-line region to a sparser, continuum-dominated region to an extremely strong absorption line like the H$\alpha$ feature.

\begin{figure}
\centering
\includegraphics[width=4in]{51peg_planet}
\caption{Phased orbit of the Hot Jupiter 51 Peg b as recovered from the \wobble\ \RVs (left) and from the \HARPS standard pipeline. While the phase coverage of \HARPS observations is sparse, we nonetheless recover orbital parameters consistent with those found by other \RV campaigns. Best-fit Keplerians for each data set are overplotted.}
\label{fig:51peg_planet}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=5.3in]{51peg_multispectrum}
\caption{Three example echelle orders at a randomly chosen epoch from the \HARPS observations of 51 Pegasi. Data are shown as dark circles, while the \wobble predictions for the stellar and telluric spectral contributions at this epoch are red and blue lines, respectively. Even in the presence of a strong stellar absorption feature such as the H$\alpha$ line (seen at 6561.5 \ang due to the star's Doppler shift at the plotted epoch), small telluric features are clearly recovered. Residuals after subtracting both star and telluric models are shown below each spectrum.}
\label{fig:51peg_spectrum}
\end{figure}

@@

Next, we tested \wobble with 237 epochs of \Mdwarf spectra. 
\Mdwarf is a mid-M dwarf, and as such its output is low in the optical; the typical \SNR of the spectra in the central wavelength regions of \HARPS ranges from 20-50. 
In practice, this means that the 7 bluest echelle orders were consistently masked and dropped from the fit in accordance with the \SNR criteria outlined previously. 
Regardless, a large amount of data remained, allowing us to test \wobble's performance on observations with substantially different properties from the previous case.

@@

\begin{figure}
\centering
\includegraphics[width=5in]{barnards_rvs}
\caption{Radial velocity measurements for \Mdwarf from \wobble (black circles) and from the standard \HARPS pipeline (purple squares). The median \RV has been subtracted from each data set. The predicted secular change in \RV due to \Mdwarf's projected motion, calculated using \gaia properties, is shown as a solid blue line (top panel). Residuals away from the predicted trend are plotted in the lower panel.}
\label{fig:barnards_rvs}
\end{figure}

@@

@@

\begin{figure}
\centering
\includegraphics[width=5.5in]{barnards_multispectrum}
\caption{Example echelle orders at a randomly chosen epoch from the \HARPS observations of Barnard's Star, plotted as in Figure \ref{fig:51peg_spectrum}.}
\label{fig:barnards_spectrum}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=6in]{barnards_model}
\caption{A subset of the \Mdwarf spectrum including the sodium doublet at 4456-7 \ang. Data for the highest-\SNR single observation are shown as black points, while the spectral template derived by \wobble from the full data set is shown as a red line. The theoretical \PHOENIX model for an M dwarf with \teff = 3200 K, \logg = 5.0, and \mh = $-0.5$ is shown as a purple line. The \wobble model and data have been Doppler-shifted by eye to match the rest-frame line positions, since \wobble does not deliver an absolute \RV. The data-driven \wobble model of the stellar spectrum shares general strong line locations and relative strengths with the \PHOENIX model, but in detail it delivers a much more accurate fit to the data than the theoretical M dwarf model achieves.}
\label{fig:barnards_model}
\end{figure}

@@

Despite the extreme noise at the bluer end of the wavelength range, where \Mdwarf is very faint, \wobble successfully retrieves a template spectrum that appears consistent with general expectations for a mid-M dwarf through most of the \HARPS wavelength range. 
As a sanity check, we compare the \wobble stellar template model with a high-resolution \PHOENIX model at \Mdwarf's previously measured spectral parameters \citep[\teff = 3200 K, \logg = 5.0, and \mh = $-0.5$;][and references therein]{Husser2013, Artigau2018}. 
Even in the bluest regions, the placements and relative strengths of the large-amplitude features inferred by \wobble generally compare well with the \PHOENIX predictions (Figure \ref{fig:barnards_model}). 
The \wobble model for \Mdwarf diverges from the theoretical models in the smaller absorption lines and in the degree of broadening for strong lines.  
\todo{(TO DO: are these known model weaknesses? ask Travis Barman about opacity binning affecting the detailed features)}

@@

\begin{figure}
\centering
\includegraphics[width=5in]{hd189_rm}
\caption{\RVs measured by \wobble for a single night of observations of the Hot 
         Jupiter host HD 189733. The \RM is clearly seen as the planet transits 
         the star. The top panel shows the data (black points), the maximum
         likelihood fit using \starry (black curve), and 500 posterior samples
         (orange lines). The bottom panel shows the residuals of the maximum
         likelihood model fit.}
\label{fig:hd189_rm}
\end{figure}

@@

These results confirm that the \wobble method measures line asymmetries as \RV shifts in the same manner as traditional \RV analysis techniques. 
While this is a useful approximation in the case of the \RM, in general the confusion of these two spectral changes is a major cause of correlated noise in \RV time series \citep[e.g.][]{}. \todo{citep}
The simplicity and extensibility of the \wobble method is useful in this regard, as the model could be modified to fit line asymmetries and true Doppler shifts as separate parameters. 
This would help to disentangle signals caused by photospheric features like starspots from the \RV signal due to reflex motion in a planetary system. 
We discuss this prospect further in Section \ref{s:future}.

@@

\begin{figure}
\centering
\includegraphics[width=5in]{telluric_standard}
\caption{Comparison of telluric spectrum derived by \wobble analysis of 51 Peg spectra (upper panel) and the spectrum observed with a telluric standard star (lower panel) around an H$_2$O absorption band. No telluric standard star data was used in determining the \wobble telluric template. The \wobble method delivers a very high \SNR telluric spectrum with no overhead observing time required.}
\label{fig:telluric_standard}
\end{figure}

>>>>

\documentclass[twocolumn]{aastex62}

@@

\begin{abstract}
% Context
Extreme-precision radial velocity (\EPRV) exoplanet surveys face considerable data analysis challenges in extracting maximally precise \RVs from spectra. 
Chief among these, particularly for the upcoming generation of red-optimized spectrographs targeting M dwarfs, is the presence of telluric absorption features which are not perfectly known. 
Another major limitation on the achievable \RV precision is the need to adopt an imperfect stellar template against which to cross-correlate or otherwise match the observations. 
In both cases, this precision-limiting reliance on external information can be sidestepped using the data itself. 
% Aims
Here we propose a data-driven method to simultaneously extract precise \RVs and infer the underlying stellar and telluric spectra using a linear model (in the log of flux). 
The model employs a convex objective and convex regularization to keep the optimization of the spectral components fast. 
We implement this method in \wobble, an open-source \python package which uses \TF in one of its first non-neural-network applications to astronomical data. 
In this work, we demonstrate the performance of \wobble on archival \HARPS spectra.
% Results
We recover the canonical exoplanet 51 Pegasi b, detect the secular \RV evolution of the M dwarf \Mdwarf, and retrieve the \RM for the Hot Jupiter HD 189733b. 
We also present extremely high-\SNR composite stellar spectra and detailed time-variable telluric spectra derived from these data. 
\end{abstract}

@@

Data analysis pipelines for absorption cell instruments have long used a stellar template that is customized to the star in question, although traditionally this template is a single observation taken for the purpose \citep[e.g.][]{Butler1996}. 
Higher quality templates can be derived iteratively from the data \citep{Sato2002, Gao2016}. 
Such data-driven templates have the benefit of being both customized to the star in question and high-\SNR without requiring observing overhead.

Common to the issues of telluric correction and stellar template building is the fact that all of the necessary information about the unknown spectra is encoded in the data. 
With a simple data-driven model, one could learn both stellar and telluric spectra simultaneously with the stellar \RVs.

@@

Here we develop and implement such a data-driven model to infer the telluric and stellar spectra and calculate the stellar \RV at each observed epoch. 
The telluric model component may vary with time in a low-dimensional manner, which is also inferred from the data. 
Our model requires no prior knowledge of spectral features for the star or for the Earth's atmosphere. 
As such, it does not yield absolute measurements of \RVs, only highly precise relative measurements between epochs.

@@

In Section \ref{s:methods}, we outline the model and its key underlying assumptions. 
We present an open-source implementation of this method in \python and \TF called \wobble.
In Section \ref{s:results}, we demonstrate \wobble's capabilities by applying our method to \HARPS archival data for three target stars: the canonical planet-hosting solar analog 51 Peg, the quiet M dwarf \Mdwarf, and the Hot Jupiter host HD 189733. 
We look further into the detailed time-variable telluric spectra inferred from these data in Section \ref{s:tellurics}. 
We revisit many of the assumptions underlying \wobble in Section \ref{s:future} and outline potential ways of adapting \wobble\ for such cases as instruments with absorption cells, intrinsic time variability in the stellar spectrum, and lower-quality data. 
Finally, we conclude with a brief summary in Section \ref{s:conclusion}.

@@

To make our method practically feasible, we assume that the number of observations $N$ in the data set is large ($N \gg N_{components}$) and spread out across the observing season. 
The quantity of spectra needed is a fundamental restriction rooted in the fact that every spectrum being modeled introduces a large number of free parameters to the model. 
In most cases, the epochs of these observations will also need to span a significant fraction of the observing season. 
In order to disentangle the stellar and telluric spectral components, the spectra must undergo Doppler shifts with respect to each other that are at least as large as a resolution element of the spectrograph. 
For stars which do not physically undergo large \RV shifts over very short timescales, this means that it is necessary to observe the star over a significant fraction of the year to take advantage of the changes in the Earth's projected motion. 
We will comment on ways to overcome these restrictions in Section \ref{s:future}.

@@

We take the data to be the $M \times N$ matrix Y, where each entry $y_{m,n}$ is the observed logarithmic flux $\ln f$ for pixel $m$ of $M$ at epoch $n$ of $N$. orange
We also have a corresponding $M \times N$ matrix of wavelength solutions. 
Again, the entries of this matrix are in logarithmic form ($\ln \lambda$). 
This choice makes the Doppler shift an additive term, considerably simplifying the math to follow.

@@

The stellar spectrum contribution is: 
\begin{equation}
y_{n}^\star = P(v^{obs}_{n}) \mu^{\star},
\label{eqn:star}
\end{equation}
where $\mu^{\star}$ is a spectral template with length $M'$ corresponding to
the length of an arbitrarily-chosen wavelength grid of control points, and
$P$ is a linear operator which applies a Doppler shift by velocity $v_n$ and
interpolates $\mu^\star$ into the observatory rest frame.
Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
\begin{equation}
 \ln \lambda(v) = \ln \lambda_{0} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
\end{equation}
The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space.
$P$ could be represented as a rectangular $M \times M'$ sparse matrix, but the dot product can be implemented efficiently using linear interpolation instead of instantiating the full matrix.

The apparent stellar \RV, $v^{obs}_{n}$, is a sum of the star's actual velocity about its center of mass ($v^{\star}_{n}$) and the projected motion of the Earth about the Solar System barycenter (Barycentric Earth Radial Velocity or \BERV), the latter of which is known.

The telluric spectrum contribution is:
\begin{equation}
y_{n}^t =  A_n(\mu^{t} + W^{t} z_n).
\end{equation}
In addition to its mean spectrum $\mu^t$, the telluric component also includes a time-dependent term assembled from two variables: $W^t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$. 
$W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of basis vectors used. 
For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications. 
The net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $A_n$, a known quantity.

@@

L1 normalization adds a term to the log-likelihood that takes the form:
\begin{equation}
\Lone{\lambda}{p} \equiv -\lambda \sum_{i} | p_{i} | ,
\end{equation}
where $p$ is the vector of parameters to be normalized (in this case $\mu^{\star}$, $\mu^{t}$, or $W^{t}$) and $\lambda$ is the regularization amplitude.

@@

With regularization included, the final model likelihood to be optimized is:
\begin{equation}
\begin{split}
\ln \mathcal{L} = & -\frac{1}{2} \sum_{n} (y_{n} - y'_{n})^T C_{n}^{-1} (y_{n}-y'_{n})  \\
 & + \Lone{\lambda_1}{\mu^{\star}} + \Ltwo{\lambda_2}{\mu^{\star}}   + \Lone{\lambda_3}{\mu^{t}} + \Ltwo{\lambda_4}{\mu^{t}} \\
 &  + \Lone{\lambda_5}{W^t}  + \Ltwo{\lambda_6}{W^t} + \Lone{1.0}{z^t} ,
\end{split}
\end{equation}
%\todo{BEDELL: THIS IS WRONG. We were supposed to set $\lambda_5$ to unity. My (Hogg's) bad. However, I wrote the following to justify our WRONG choice!}
where the regularization amplitude on the basis weights is arbitrarily set to unity, and all other regularization amplitudes $\lambda$ are set by grid searches using the above-described validation procedure.
The basis-weight regularization amplitude can be set arbitrarily to unity because there is a perfect degeneracy between amplifying the basis vectors and attenuating the basis weights.
This regularization choice breaks that perfect degeneracy.
The basis weights are regularized to encourage our desired outcome where the mean telluric spectrum contains as much telluric information as possible, while the variable basis picks up only the necessary time-variable changes.

@@

As an initial guess, we set the star to be stationary, e.g. $v^{obs} = $ \BERV at all epochs. 
We may then initialize the stellar spectrum $\mu^{\star}$ by Doppler-shifting the data and calculating the median flux across \BERV-corrected spectra in bins at each model wavelength $\lambda'$. 
The telluric template is initialized similarly by using the residuals after the stellar contribution has been removed, again binning in model wavelength (this time without applying any Doppler shift to the spectra) and taking the median values of each bin. 
Finally, we initialize the telluric spectrum's basis vectors $W_t$ and weights $z_t$ by performing Principal Component Analysis (\acronym{PCA}) on the residuals after both the stellar spectrum and the mean telluric spectrum have been removed. 
The $K$ highest eigenweights and their corresponding eigenvectors are taken as the basis weights and vectors.

Once all parameters are initialized, we optimize them iteratively. 
The likelihood function is maximized first by varying the stellar and telluric templates (including the telluric basis vectors). 
This step is a convex optimization, meaning that a global optimum for the template parameters should be reached under the condition of fixed time-dependent variables. 
Next, the templates are held fixed and the time-dependent parameters (stellar \RVs and telluric basis weights) are varied to maximize the likelihood function again. 
Technically the velocities are location parameters and their optimization is not convex.
However, once the velocities are known to a small fraction of a pixel (which they usually are, in practice, in advance of running \wobble), the linearized problem becomes convex at each iteration.
We repeat this procedure, optimizing spectra and time-dependent parameters in turn, until the likelihood appears to converge, typically within 100 iterations.
This iterative procedure is equivalent to a full simultaneous optimization in the limit of many iterations to convergence.

@@

The above-described model can be implemented in a variety of ways. We chose to build our code, \wobble, in \code{python} using \TF \citep{Abadi15}. 
\TF is a model building framework that has been primarily designed for machine
learning applications, but at its core \TF is fundamentally a collection of
highly optimized routines for doing linear algebra and efficiently computing the
derivatives of these models with respect to large numbers of parameters.
\wobble can be represented in this framework so we have re-purposed \TF for our
needs.
By using this framework, we benefit from the high performance and scalability
of the implementation, as well as the algorithms implemented within \TF for
fitting large numbers of parameters to large datasets.
%While it has primarily been used in the context of deep learning in the astronomical literature, its auto-differentiation and fast linear algebra capabilities make it valuable for this application as well.
The necessary optimizations over many parameters can be performed with high efficiency by \TF: the below-described analysis of 91 \HARPS spectra of 51 Peg, including the optimization of 72 spectral orders, runs in 60 minutes on a standard Mac desktop.

@@

\begin{figure*}[ht!]
\centering
\includegraphics[width=5in]{51peg_planet}
\caption{Phased orbit of the Hot Jupiter 51 Peg b as recovered from the \wobble\ \RVs (left) and from the \HARPS standard pipeline. While the phase coverage of \HARPS observations is sparse, we nonetheless recover orbital parameters consistent with those found by other \RV campaigns. Maximum-likelihood Keplerians (solid orange lines) and 1-$\sigma$ posterior constraints (shaded orange regions) for each data set are overplotted.}
\label{fig:51peg_planet}
\end{figure*}

\begin{deluxetable*}{cccc}
\tablehead{
\colhead{Parameter} & \colhead{Units} & \colhead{\wobble Fit} & \colhead{\HARPS Pipeline Fit}}
\tablecaption{Orbital Parameters of 51 Pegasi b}
\label{tbl:51peg}
\startdata
$K$ & m s$^{-1}$ & $55.57^{+2.28}_{-2.04}$ & $53.84^{+1.96}_{-2.20}$ \\
$P$ & days & $4.2292^{+0.0003}_{-0.0003}$ & $4.2294^{+0.0003}_{-0.0003}$ \\
$t_0$ & JD & $2456546.89^{+0.02}_{-0.02}$ & $2456546.94^{+0.01}_{-0.02}$ \\
$e$ & $-$  & $0.03^{+0.02}_{-0.02}$ & $0.04^{+0.02}_{-0.02}$ \\
$\omega$ & rad & $0.45^{+0.62}_{-1.00}$ & $-1.79^{+0.34}_{-0.32}$ \\
$s$ & m s$^{-1}$ & $0.01^{+0.08}_{-0.01}$ & $0.74^{+0.06}_{-0.07}$ \\
$m_p$ sin$i$ & $M_J$ & $0.41^{+0.02}_{-0.02}$ & $0.40^{+0.01}_{-0.02}$ \\
\enddata
\end{deluxetable*}

@@

Despite the sparse phase coverage available across the planetary orbit, we recover a signal at the expected period and semi-amplitude in the \RVs (Figure \ref{fig:51peg_planet}). 
We fit a Keplerian signal with seven free parameters (period $P$, \RV semi-amplitude $K$, eccentricity $e$, argument of periastron $\omega$, time of periastron $T_0$, \RV offset $c$, and \RV jitter $s$) using the \code{exoplanet} package \citep{exoplanet}. 
The resulting best-fit parameters for the \wobble\ \RVs are generally consistent with expectations based on literature values (Table \ref{tbl:51peg}). 
The \RV uncertainties derived by \wobble appear accurate based on the negligible jitter in the fit, while the photon-noise-based \RV errors provided by the \HARPS pipeline are much smaller and require a significant jitter to achieve a good fit.

\begin{figure*}[t]
\centering
\includegraphics[width=5.3in]{51peg_multispectrum}
\caption{Three example echelle orders at a randomly chosen epoch from the \HARPS observations of 51 Pegasi. Data are shown as dark circles, while the \wobble predictions for the stellar and telluric spectral contributions at this epoch are red and blue lines, respectively. Even in the presence of a strong stellar absorption feature such as the H$\alpha$ line (seen at 6561.5 \ang due to the star's Doppler shift at the plotted epoch), small telluric features are clearly recovered. Residuals after subtracting both star and telluric models are shown below each spectrum.}
\label{fig:51peg_spectrum}
\end{figure*}

Fits to an individual spectrum are shown in Figure \ref{fig:51peg_spectrum}. 
We emphasize that no \textit{a priori} information on e.g. expected spectral line positions and depths were used. 
The \wobble algorithm as it is currently implemented treats each control point of the template spectra as independent and has no line shape parameterization included. 
Nevertheless, the optimized stellar templates clearly reproduce the expected appearance of the spectra in a variety of regimes, from a crowded-line region to a sparser, continuum-dominated region to an extremely strong absorption line like the H$\alpha$ feature. 
Moreover, telluric absorption features are recovered down to a very small amplitude.

@@

\begin{figure*}
\centering
\includegraphics[width=5in]{barnards_rvs}
\caption{Radial velocity measurements for \Mdwarf from \wobble (black circles) and from the standard \HARPS pipeline (blue squares). The median \RV has been subtracted from each data set. The predicted secular change in \RV due to \Mdwarf's projected motion, calculated using \gaia properties, is shown as a solid orange line (top panel). Residuals away from the predicted trend are plotted in the lower panel.}
\label{fig:barnards_rvs}
\end{figure*}

Next, we tested \wobble with 237 epochs of \Mdwarf spectra. 
We did not utilize the additional spectra taken after the \HARPS 2015 optical fiber upgrade, as these would need to be treated as an independent data set \citep{LoCurto2015}. 
\Mdwarf is a mid-M dwarf, and as such its output is low in the optical; the typical \SNR of the spectra in the central wavelength regions of \HARPS ranges from 20-50. 
In practice, this means that the 7 bluest echelle orders were consistently masked and dropped from the fit in accordance with the \SNR criteria outlined previously. 
Regardless, a large amount of data remained, allowing us to test \wobble's performance on observations with substantially different properties from the previous case.

@@

@@

\begin{figure*}[ht!]
\centering
\includegraphics[width=5.5in]{barnards_multispectrum}
\caption{Example echelle orders at a randomly chosen epoch from the \HARPS observations of Barnard's Star, plotted as in Figure \ref{fig:51peg_spectrum}.}
\label{fig:barnards_spectrum}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=6in]{barnards_model}
\caption{A subset of the \Mdwarf spectrum including the sodium doublet at 4456-7 \ang. Data for the highest-\SNR single observation are shown as black points, while the spectral template derived by \wobble from the full data set is shown as a red line. The theoretical \PHOENIX model for an M dwarf with \teff = 3200 K, \logg = 5.0, and \mh = $-0.5$ is shown as an orange line. The \wobble model and data have been Doppler-shifted by eye to match the rest-frame line positions, since \wobble does not deliver an absolute \RV. The data-driven \wobble model of the stellar spectrum shares general strong line locations and relative strengths with the \PHOENIX model, but in detail it delivers a much more accurate fit to the data than the theoretical M dwarf model achieves.}
\label{fig:barnards_model}
\end{figure*}

@@

@@

Despite the extreme noise at the bluer end of the wavelength range, where \Mdwarf is very faint, \wobble successfully retrieves a template spectrum that appears consistent with general expectations for a mid-M dwarf through most of the \HARPS wavelength range. 
As a sanity check, we compare the \wobble stellar template model with a high-resolution \PHOENIX model at \Mdwarf's previously measured spectral parameters \citep[\teff = 3200 K, \logg = 5.0, and \mh = $-0.5$;][and references therein]{Husser2013, Artigau2018}. 
Even in the bluest regions, the placements and relative strengths of the large-amplitude features inferred by \wobble generally compare well with the \PHOENIX predictions (Figure \ref{fig:barnards_model}). 
The \wobble model for \Mdwarf diverges from the theoretical models in the smaller absorption lines and in the degree of broadening for strong lines.  
%\todo{(TO DO: are these known model weaknesses? ask Travis Barman about opacity binning affecting the detailed features)}

@@

\begin{figure*}
\centering
\includegraphics[width=5in]{hd189_rm}
\caption{\RVs measured by \wobble for a single night of observations of the Hot 
         Jupiter host HD 189733. The \RM is clearly seen as the planet transits 
         the star. The top panel shows the data (black points), the maximum
         likelihood fit using \starry (black curve), and 500 posterior samples
         (orange lines). The bottom panel shows the residuals of the maximum
         likelihood model fit.}
\label{fig:hd189_rm}
\end{figure*}

@@

These results confirm that the \wobble method measures line asymmetries as \RV shifts in the same manner as traditional \RV analysis techniques. 
While this is a useful approximation in the case of the \RM, in general the confusion of these two spectral changes is a major cause of correlated noise in \RV time series \citep[e.g.][]{Queloz2001}. 
The simplicity and extensibility of the \wobble method is useful in this regard, as the model could be modified to fit line asymmetries and true Doppler shifts as separate parameters. 
This would help to disentangle signals caused by photospheric features like starspots from the \RV signal due to reflex motion in a planetary system. 
We discuss this prospect further in Section \ref{s:future}.

@@

\begin{figure*}
\centering
\includegraphics[width=5in]{telluric_standard}
\caption{Comparison of telluric spectrum derived by \wobble analysis of 51 Peg spectra (upper panel) and the spectrum observed with a telluric standard star (lower panel) around an H$_2$O absorption band. No telluric standard star data was used in determining the \wobble telluric template. The \wobble method delivers a very high \SNR telluric spectrum with no overhead observing time required.}
\label{fig:telluric_standard}
\end{figure*}
